---
title: "Getting Started with multiGGM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with multiGGM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The **multiGGM** package implements the Bayesian method of Peterson, Stingo
& Vannucci (2015, *JASA*) for joint inference of multiple Gaussian graphical
models (GGMs). Given data from $K$ groups, the method simultaneously
estimates group-specific precision matrices $\Omega_k$ and their
associated conditional independence graphs $G_k$, while borrowing
strength across groups through a Markov random field (MRF) prior.

This vignette walks through:

1. Simulating multi-group data with known graph structure
2. Fitting the model with `multiggm_mcmc()`
3. Checking convergence
4. Extracting and interpreting results
5. Comparing estimated graphs to ground truth

## 1. Simulate data

We generate data from two groups with $p = 15$ variables and $n = 100$
observations each, using a banded (AR-2) graph structure.

```{r simulate}
library(multiGGM)

sim <- simulate_multiggm(
  K = 2, p = 15, n = 100,
  graph_type = "band",
  perturb_prob = 0.1,
  seed = 42
)

cat("True edge counts:\n")
for (k in 1:sim$K) {
  adj_k <- sim$adj_list[[k]]
  cat(sprintf("  Group %d: %d edges\n", k, sum(adj_k[upper.tri(adj_k)])))
}
```

## 2. Fit the model

We pass raw data matrices directly. The function computes cross-product
matrices internally after column-centering.

```{r fit, results='hide'}
fit <- multiggm_mcmc(
  data_list = sim$data_list,
  burnin = 2000,
  nsave = 1000,
  thin = 1,
  seed = 123
)
```

```{r print-fit}
fit
```

## 3. Check convergence

### Summary

The `summary()` method reports acceptance rates and edge counts.
Acceptance rates between 20-60% are generally healthy.

```{r summary}
summary(fit)
```

### Trace plots

Trace plots help assess mixing. The theta trace shows the graph similarity
parameter -- values above zero indicate the model is borrowing strength
across groups.

```{r trace-theta, fig.height=3}
plot(fit, type = "trace_theta")
```

```{r trace-edges, fig.height=3}
plot(fit, type = "trace_edges")
```

## 4. Extract results

### Posterior inclusion probabilities (PIP)

PIPs indicate how often each edge was included across posterior samples.
Higher PIP = stronger evidence for that edge.

```{r pip}
pip <- pip_edges(fit)
```

```{r pip-heatmap}
plot(fit, type = "pip")
```

### Estimated precision and partial correlations

The `coef()` method returns posterior mean precision matrices, while
`fitted()` returns posterior mean partial correlations.

```{r coef}
omega_hat <- coef(fit)
cat("Posterior mean precision (Group 1, first 5x5 block):\n")
round(omega_hat[[1]][1:5, 1:5], 3)
```

```{r fitted}
pcor_hat <- fitted(fit)
cat("Posterior mean partial correlations (Group 1, first 5x5 block):\n")
round(pcor_hat[[1]][1:5, 1:5], 3)
```

### Credible intervals

```{r ci}
pcor_draws <- posterior_pcor(fit)
ci <- posterior_ci(pcor_draws)

# Example: 95% CI for edge (1,2) in group 1
cat(sprintf("pcor[1,2] Group 1: %.3f (%.3f, %.3f)\n",
            ci$median[1, 2, 1], ci$lower[1, 2, 1], ci$upper[1, 2, 1]))
```

## 5. Compare to ground truth

### Visual comparison

`plot_recovery()` shows the true graph, PIP heatmap, and thresholded
estimate side-by-side.

```{r recovery, fig.width=9, fig.height=4}
cm <- plot_recovery(fit, sim$adj_list, pip_threshold = 0.5)
```

### Confusion metrics

```{r confusion}
for (k in seq_along(cm)) {
  cat(sprintf("Group %d: TP=%d, FP=%d, FN=%d, TN=%d, TPR=%.2f, FPR=%.2f\n",
              k, cm[[k]]["TP"], cm[[k]]["FP"], cm[[k]]["FN"], cm[[k]]["TN"],
              cm[[k]]["TPR"], cm[[k]]["FPR"]))
}
```

### ROC curve

The ROC curve evaluates edge selection across all possible thresholds.

```{r roc, fig.height=4, fig.width=4.5}
roc1 <- roc_auc(pip[, , 1], sim$adj_list[[1]])
plot_roc(roc1, main = "ROC - Group 1")
```

## 6. Network visualization

We can visualize the estimated graph as a network.

```{r network, fig.height=5, fig.width=10}
par(mfrow = c(1, 2))
plot_network(sim$adj_list[[1]], main = "True - Group 1", vertex_size = 12)
est_adj <- (pip[, , 1] >= 0.5) * 1L
plot_network(est_adj, main = "Estimated - Group 1", vertex_size = 12)
```

## 7. Differential edges

For $K = 2$ groups, we can identify edges that differ between groups.

```{r diff-edges}
pip_diff <- pip_diff_edge(fit)
cat("Edges with P(differential) > 0.5:",
    sum(pip_diff[upper.tri(pip_diff)] > 0.5), "\n")
```

```{r diff-pcor}
pcor_draws <- posterior_pcor(fit)
diff_prob <- diff_prob_pcor(pcor_draws, delta = 0.1)
cat("Edges with P(|pcor diff| > 0.1) > 0.5:",
    sum(diff_prob[upper.tri(diff_prob)] > 0.5), "\n")
```

## References

Peterson, C.B., Stingo, F.C. & Vannucci, M. (2015). Bayesian inference
of multiple Gaussian graphical models. *Journal of the American
Statistical Association*, 110(509), 159-174.
