---
title: "Scalable Inference with the SSVS Method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scalable Inference with the SSVS Method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

Version 0.2.0 of **multiGGM** adds a second inference method: the
spike-and-slab continuous shrinkage prior of Shaddox et al. (2018,
*Statistics in Biosciences*). This method replaces the G-Wishart prior
and Wang-Li exchange algorithm with a column-wise Gibbs sampler that
scales to graphs with $p > 100$ nodes.

Select the method via the `method` argument:

```r
# Original G-Wishart method (default, best for small p)
fit_gw <- multiggm_mcmc(data_list, method = "gwishart")

# Scalable SSVS method (for larger p)
fit_ssvs <- multiggm_mcmc(data_list, method = "ssvs")
```

Both methods estimate the same quantities---group-specific precision
matrices $\Omega_k$, adjacency matrices $G_k$, graph similarity
parameters $\theta_{km}$, and edge-specific log-odds $\nu_{ij}$---and
the output objects have an identical structure. All post-processing
functions (`pip_edges`, `posterior_pcor`, `plot`, etc.) work with either
method.

## When to use each method

| Scenario | Recommended method |
|----------|-------------------|
| $p \le 30$, need exact G-Wishart inference | `"gwishart"` |
| $p > 30$, exploratory analysis | `"ssvs"` |
| $p > 50$, any use case | `"ssvs"` (G-Wishart becomes prohibitively slow) |
| Comparing both priors on the same data | Run both, compare PIPs |

## Quick start example

```{r ssvs-fit, results='hide'}
library(multiGGM)

sim <- simulate_multiggm(K = 2, p = 30, n = 100, seed = 42)

fit <- multiggm_mcmc(
  data_list = sim$data_list,
  method    = "ssvs",
  burnin    = 2000,
  nsave     = 1000,
  seed      = 123
)

# TODO add section on multiple chains, how to use them and how to evaluate convergence, and when it is recommended to use more than one chain
```

```{r ssvs-summary}
summary(fit, pip_threshold = 0.5)
```

```{r ssvs-pip, fig.height=4}
plot(fit, type = "pip")
```

```{r ssvs-trace, fig.height=3}
plot(fit, type = "trace_theta")
```

## Scalability benchmarks

The table below shows wall-clock times for 200 burn-in + 100 saved
iterations with $K = 2$ groups and $n = 100$ observations per group,
measured on an Apple M1 processor.

| $p$ | SSVS | G-Wishart | Speedup |
|-----|------|-----------|---------|
| 10  | 0.1 s | 0.8 s | 6x |
| 20  | 0.7 s | 10.9 s | 17x |
| 30  | 1.7 s | 104.9 s | 63x |
| 50  | 7.5 s | (infeasible) | --- |

The speedup increases with $p$ because the G-Wishart method's cost grows
as $O(K \cdot p^4)$ per iteration (due to per-edge exchange proposals and
BIPS clique updates), while the SSVS method grows as $O(K \cdot p^3)$
(column-wise Gibbs sweeps with rank-1 covariance updates).

## How the two methods differ

Both methods share the same Markov random field (MRF) prior that
borrows strength across groups, and the same Metropolis-Hastings updates
for the graph similarity parameters $\theta_{km}$ and edge-specific
log-odds $\nu_{ij}$. They differ only in how the precision matrices
$\Omega_k$ and graphs $G_k$ are sampled.

### G-Wishart method (`method = "gwishart"`)

The original Peterson et al. (2015) approach places a G-Wishart prior
$\Omega_k \mid G_k \sim W_G(b, D)$ on each precision matrix. Sampling
proceeds via the Wang-Li (2012) exchange algorithm:

- **For each edge** $(i,j)$ in each group $k$:
  1. Compute a Bernoulli proposal probability from the G-Wishart
     posterior ratio (`log_H`) and the MRF similarity term.
  2. If the proposed graph differs, sample a new precision matrix from
     the G-Wishart conditional (`GWishart_NOij_Gibbs`).
  3. Accept or reject via an exchange ratio involving G-Wishart pdf
     evaluations (`log_GWishart_NOij_pdf`).
  4. Update the precision matrix from the posterior conditional.
- **After all edges**: refresh the full precision matrix via BIPS
  (Block Incomplete Partial correlation Sampling) using maximal clique
  decomposition (`GWishart_BIPS_maximumClique`).

This requires $O(p^2)$ edge proposals per group, each involving
$O(p^2)$--$O(p^3)$ matrix operations, plus an additional $O(p^2)$
BIPS step.

### SSVS method (`method = "ssvs"`)

The Shaddox et al. (2018) approach replaces the G-Wishart prior with a
spike-and-slab normal prior on each off-diagonal precision element:

$$\omega_{k,ij} \mid z_{k,ij} \sim (1 - z_{k,ij}) \cdot N(0, v_0) + z_{k,ij} \cdot N(0, v_1)$$

where $v_0 \ll v_1$ (spike vs. slab), and an exponential prior on diagonal
elements: $\omega_{k,ii} \sim \text{Exp}(\lambda/2)$. Because this prior
has a tractable normalizing constant, sampling uses a direct column-wise
Gibbs sweep:

- **For each column** $i$ of the precision matrix in each group $k$:
  1. Compute the Schur complement
     $C_{11}^{-1} = \Sigma_{11} - \Sigma_{12}\Sigma_{12}^\top / \Sigma_{ii}$
     from the current covariance (no matrix inversion needed).
  2. Form the posterior precision
     $C_i = (S_{ii} + \lambda) \cdot C_{11}^{-1} + \text{diag}(1/\tau)$,
     where $\tau$ contains $v_0$ or $v_1$ depending on current edge
     inclusion.
  3. Sample the off-diagonal column from its multivariate normal
     conditional: $\epsilon \sim N(\mu_i, C_i^{-1})$.
  4. Sample the diagonal from a Gamma conditional:
     $\gamma \sim \text{Gamma}(n_k/2 + 1,\; (S_{ii} + \lambda)/2)$.
  5. **Rank-1 covariance update**: update $\Sigma$ directly from the
     new column without any full matrix inversion.
  6. **Update edge indicators** $z$ via posterior spike-slab odds,
     incorporating the MRF prior analytically through the conditional
     inclusion probability.

There is no exchange algorithm, no BIPS clique step, and no
trans-dimensional moves. Each iteration costs $O(p)$ column sweeps, each
$O(p^2)$ for the Cholesky solve and rank-1 update, giving $O(p^3)$ total
per group per iteration.

### What stays the same

- The **MRF normalizing constant** $C(\theta, \nu_{ij})$ uses the same
  exact enumeration over $\{0,1\}^K$ binary vectors (feasible when $K$
  is small).
- The **$\theta_{km}$ update** (between-model spike/slab toggle +
  within-model Gamma proposal) and the **$\nu_{ij}$ update** (MH with
  Beta proposal) are identical in both methods.

### Key differences at a glance

| Aspect | G-Wishart | SSVS |
|--------|-----------|------|
| **Precision prior** | G-Wishart $W_G(b, D)$ | Spike-slab $N(0, v_0)$ / $N(0, v_1)$ |
| **Graph sampling** | Exchange algorithm (MH per edge) | Posterior Bernoulli from spike-slab odds |
| **Precision sampling** | NOij Gibbs per edge + BIPS clique | Column-wise Gaussian conditional |
| **Covariance updates** | Full inversion after BIPS | Rank-1 incremental update |
| **Complexity/iter** | $O(K \cdot p^4)$ | $O(K \cdot p^3)$ |
| **Scalability** | $p \lesssim 30\text{--}50$ | $p > 100$ demonstrated |

## SSVS hyperparameters

### Default values

The SSVS-specific hyperparameters and their defaults are:

| Parameter | Default | Role |
|-----------|---------|------|
| `v0` | $0.02^2 = 0.0004$ | Spike variance (small, effectively zeroes the edge) |
| `v1` | $50^2 \times v_0 = 1.0$ | Slab variance (large, allows substantial precision values) |
| `lambda` | 1 | Exponential rate for diagonal elements |

The variance inflation ratio $h = v_1 / v_0 = 2500$ controls edge
selection sensitivity. The prior edge inclusion probability adapts to
the dimension: $\pi = 2/(p-1)$, which favors sparse graphs as $p$ grows.

These defaults follow Wang (2015, *Bayesian Analysis*), "Scaling it up:
stochastic search structure learning in graphical models," which
introduced the SSVS approach for single GGMs. The multiGGM
implementation extends it to multiple graphs by adding MRF-coupled edge
indicators.

### Common hyperparameters (shared with G-Wishart)

| Parameter | Default | Role |
|-----------|---------|------|
| `a`, `b` | 1, 4 | Beta(1, 4) prior on edge inclusion log-odds $\nu_{ij}$; gives prior $P(\text{edge}) \approx 0.20$ |
| `alpha`, `beta` | 2, 5 | Gamma(2, rate=5) slab prior on $\theta_{km}$; mean = 0.4 |
| `w` | 0.9 | $P(\gamma_{km} = 1)$; strong prior belief that groups share structure |
| `alpha_prop`, `beta_prop` | 1, 1 | Gamma proposal for $\theta_{km}$ MH step (scale parameterization) |

### Tuning recommendations

- **`v0` and `v1`**: The defaults work well for standardized data. If
  your data are on a very different scale, you may need to adjust. A
  larger `v1` allows stronger partial correlations; a smaller `v0`
  provides stricter edge selection.
- **`lambda`**: Controls diagonal shrinkage. The default of 1 is
  generally safe. Increasing `lambda` adds regularization to the
  diagonal.
- **`burnin` and `nsave`**: The SSVS sampler typically mixes faster
  than G-Wishart, so you may need fewer iterations. Start with
  `burnin = 5000, nsave = 5000` and check trace plots. For production
  runs, Shaddox et al. (2018) used `burnin = 40000, nsave = 80000`.
- **`a` and `b`**: Decrease `b` for denser graphs (e.g., Beta(1, 1)
  for uniform prior on edge probability) or increase for sparser graphs
  (e.g., Beta(1, 9) for $\sim$10% prior edge probability).

### Passing custom hyperparameters

```r
fit <- multiggm_mcmc(
  data_list = sim$data_list,
  method    = "ssvs",
  hyper     = list(
    v0    = 0.01^2,    # tighter spike
    v1    = 100^2 * 0.01^2,  # wider slab
    lambda = 2,        # stronger diagonal shrinkage
    a = 1, b = 9,      # sparser prior (Beta(1,9))
    alpha = 2, beta = 5,
    w = 0.5,           # less certain groups are related
    alpha_prop = 1, beta_prop = 1
  ),
  burnin = 5000,
  nsave  = 5000,
  seed   = 42
)
```

## Comparing both methods on the same data

When $p$ is small enough to run both, comparing results provides a
useful robustness check.

```{r compare, results='hide'}
sim_small <- simulate_multiggm(K = 2, p = 15, n = 100, seed = 42, graph_type = "hub")

fit_gw <- multiggm_mcmc(
  data_list = sim_small$data_list,
  method    = "gwishart",
  burnin    = 2000,
  nsave     = 1000,
  seed      = 123
)

fit_ssvs <- multiggm_mcmc(
  data_list = sim_small$data_list,
  method    = "ssvs",
  hyper = list(
  v0    = 0.001,   
    v1    = 1,  
    lambda = 1,       
    a = 1, b = 4,    
    alpha = 2, beta = 5,
    w = 0.9,           # less certain groups are related
    alpha_prop = 1, beta_prop = 1),
  burnin    = 2000,
  nsave     = 1000,
  seed      = 123
)
```

```{r compare-pips}
pip_gw   <- pip_edges(fit_gw)
pip_ssvs <- pip_edges(fit_ssvs)

# Compare PIPs for group 1 (upper triangle only)
idx <- upper.tri(pip_gw[, , 1])
cat(sprintf("Correlation of PIPs (Group 1): %.3f\n",
            cor(pip_gw[, , 1][idx], pip_ssvs[, , 1][idx])))
cat(sprintf("G-Wishart edges (PIP > 0.5):   %d\n",
            sum(pip_gw[, , 1][idx] > 0.5)))
cat(sprintf("SSVS edges (PIP > 0.5):        %d\n",
            sum(pip_ssvs[, , 1][idx] > 0.5)))
```

## Recovery comparison

```{r compare-recovery, fig.width=9, fig.height=7}
par(mfrow = c(2, 1))
cm_gw   <- plot_recovery(fit_gw,   sim_small$adj_list, pip_threshold = 0.5)
cm_ssvs <- plot_recovery(fit_ssvs, sim_small$adj_list, pip_threshold = 0.3)
```

```{r compare-metrics}
for (method_name in c("G-Wishart", "SSVS")) {
  cm <- if (method_name == "G-Wishart") cm_gw else cm_ssvs
  for (k in seq_along(cm)) {
    cat(sprintf("%s Group %d: TPR=%.2f, FPR=%.2f\n",
                method_name, k, cm[[k]]["TPR"], cm[[k]]["FPR"]))
  }
}
```

## References

Peterson, C.B., Stingo, F.C. & Vannucci, M. (2015). Bayesian inference
of multiple Gaussian graphical models. *Journal of the American
Statistical Association*, 110(509), 159--174.

Shaddox, E., Stingo, F.C., Peterson, C.B., et al. (2018). A Bayesian
approach for learning gene networks underlying disease severity in COPD.
*Statistics in Biosciences*, 10(1), 59--85.

Wang, H. (2015). Scaling it up: stochastic search structure learning in
graphical models. *Bayesian Analysis*, 10(2), 351--377.

Wang, H. & Li, S.Z. (2012). Efficient Gaussian graphical model
determination under G-Wishart prior distributions. *Electronic Journal
of Statistics*, 6, 168--198.
